---
title: "Vignette of the package `CompSign`"
author: "Lena Morrill Gavarró, Dominique-Laurent Couturier, Florian Markowetz"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{my-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: "2023"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  cache = TRUE,
  comment = "#>"
)
```

`CompSign` is a toolkit for differential abundance analysis of mutational signatures using a mixed effects Dirichlet-multinominal model (or simpler variations). The compositional nature of mutational signature exposures has often been overlooked but has important implications, as the analyses must be done in relative terms.

## Installation 
`CompSign` can be installed as usual from github:

```{r include=TRUE, eval=FALSE}
library(devtools)
devtools::install_github("lm687/CompSign")
```

These are the libraries used in this vignette:
```{r libraries, include=TRUE, eval=TRUE, message=FALSE}
library(CompSign)
library(gridExtra)
library(TMB)
```

## Datasets
Two datasets can be loaded as follows using the `data()` function:

```{r, load_data}
data(PancEndocrine_signaturesMSE)
data(ProstAdenoCA_chrom)
data(res_patient_lambda)
```

The package contains the following datasets of exposures of mutational signatures and metadata of the corresponding samples. These datasets are:

- `PancEndocrine_signaturesMSE`: Signature exposures for early and late mutations, in the PCAWG Panc-Endocrine cohort
- `ProstAdenoCA_chrom`: Signature exposures for each chromosome, in the PCAWG Prost-AdenoCA cohort

## Example workflow
In this example workflow we determine differential abundance between clonal and subclonal mutations in Pancreatic neuroendocrine tumors.

`PancEndocrine_signaturesMSE` is an object of class `sign`

```{r loading_panc-endocrine, message=FALSE, results='hide'}
PancEndocrine_signaturesMSE = load_PCAWG("../inst/extdata/roo/Panc-Endocrine_signaturesMSE_ROO.RDS",
                                         read_directly = T,
                                         typedata = "signaturesMSE", override_warning_X_Z = T)
PancEndocrine_signaturesMSE_v2 = load_PCAWG(ct = "Panc-Endocrine", typedata = "signaturesMSE", path_to_data = "../inst/extdata/", load_all_sigs = F, override_warning_X_Z = T)
```

All samples - clonal and subclonal - sorted by increasing SBS3 exposure:
```{r panc-endocrine2, fig.height=3.5, fig.width=5, message=FALSE, results='hide'}

createBarplot(normalise_rw(non_duplicated_rows(PancEndocrine_signaturesMSE$Y)),
              order_labels = names(sort(non_duplicated_rows(PancEndocrine_signaturesMSE$Y)[,'SBS3'],
                                        decreasing = F)), remove_labels=T)+ggtitle('Sorted by SBS3')

```

We create a simplified object containing exposures of fewer signatures (i.e. a subcomposition of the original signature vectors):
```{r simplify_object}
simplified_object <- give_subset_sigs_TMBobj(PancEndocrine_signaturesMSE,
                                             sigs_to_remove = c('SBS13', 'SBS17a', 'SBS17b', 'SBS30'))
```

The clonal and subclonal exposures are, respectively, the two barplots below:
```{r simplify_object2, fig.height=3, fig.width=7, message=FALSE}
do.call('grid.arrange', list(grobs=lapply(split_matrix_in_half(simplified_object$Y), function(i) createBarplot(normalise_rw(i), remove_labels = T)), nrow=1))
```

## Running the model for differential abundance

Running the model `diagREDMsinglelambda` with the dataset `simplified_object`:

```{r panc-endocrine3, message=FALSE, echo=TRUE, results='hide'}
diagDM_no_small_sigs <- wrapper_run_TMB(object = simplified_object,
                                        model = "diagREDMsinglelambda", use_nlminb=T, smart_init_vals=F)
```

This is the resulting object with the estimates and their standard deviations:

```{r TMB_output_example}
diagDM_no_small_sigs
```

These are the betas for this model
```{r plot_betas_example, results='hide', fig.height=2.5, fig.width=5}
plot_betas(diagDM_no_small_sigs)
```

and the p-value indicating differential abundance

```{r, wald_test}
wald_TMB_wrapper(diagDM_no_small_sigs)
```

## Other models
Other models can be run as follows:

```{r, eval=FALSE, echo=TRUE}
res <- wrapper_run_TMB(object = simplified_object,
                       model = "diagREDMsinglelambda", use_nlminb=T, smart_init_vals=F)
res <- wrapper_run_TMB(object = simplified_object,
                       model = "diagRE_DM", use_nlminb=T, smart_init_vals=F)
res <- wrapper_run_TMB(object = simplified_object,
                       model = "diagRE_M", use_nlminb=T, smart_init_vals=F)
res <- wrapper_run_TMB(object = simplified_object,
                       model = "FEDMsinglelambda", use_nlminb=T, smart_init_vals=F)
res <- wrapper_run_TMB(object = simplified_object,
                       model = "FE_DM", use_nlminb=T, smart_init_vals=F)
res <- wrapper_run_TMB(object = simplified_object,
                       model = "fullREDMsinglelambda", use_nlminb=T, smart_init_vals=F)
res <- wrapper_run_TMB(object = simplified_object,
                       model = "fullRE_DMonefixedlambda", use_nlminb=T, smart_init_vals=F)
res <- wrapper_run_TMB(object = simplified_object,
                       model = "fullRE_DM", use_nlminb=T, smart_init_vals=F)
res <- wrapper_run_TMB(object = simplified_object,
                       model = "fullRE_M", use_nlminb=T, smart_init_vals=F)
res <- wrapper_run_TMB(object = simplified_object,
                       model = "singleRE_DM", use_nlminb=T, smart_init_vals=F)
```

## Patient-specific dispersion parameters
In a scenario where we have several samples for each patient, which can still be grouped in two or more groups (fixed effects), we might be interested in patient-specific dispersion parameters, besides the patient-specific intercepts that we had. This has been implemented:

```{r}
library(TMB)
library(HMP)
library(ggplot2)
theme_set(theme_bw())
library(gridExtra)
# library(CompSign)  ## replace this once this is done using the package
# source("../R/helper_functions.R") ## replace this once this is done using the package
# source("../R/DA_functions.R") ## replace this once this is done using the package

##-----------------------------------------------------------------------------------------------------##

##-----------------------------------------------------------------------------------------------------##
## replace this once this is done using the package
# TMB::compile("../src/diagREpatientlambda_ME_dirichletmultinomial.cpp",  "-std=gnu++17")
# dyn.load(dynlib("../src/diagREpatientlambda_ME_dirichletmultinomial"))
##-----------------------------------------------------------------------------------------------------##

##-----------------------------------------------------------------------------------------------------##

n <- 40 ## patients
num_samples_per_patient = 30
stopifnot((num_samples_per_patient %% 2)==0)
z <- do.call('rbind', replicate(expr = diag(n), n= num_samples_per_patient, simplify = F))
x <- cbind(rep(1, n*num_samples_per_patient),
           c(rep(0, (n*num_samples_per_patient)/2),
             rep(1, (n*num_samples_per_patient)/2)))
d = 5 ## number of signatures

random_effects <- matrix(runif(n*(d-1)), nrow=n, ncol=d-1)
fixed_effects <- matrix(runif(2*(d-1)), nrow=2, ncol=d-1)

alpha_mat <- x %*% fixed_effects + z %*% random_effects
alpha_mat_softmax = softmax( cbind(alpha_mat, 0) )
rowSums(alpha_mat_softmax)

## lambda per patient
log_lambda_vec = runif(n, 3, 4)

## get patient-specific overdispersions
log_lambda_vec_per_obs <- z %*% log_lambda_vec
log_lambda_vec_per_obs

## simulate counts
Nm = rep(1000, n*num_samples_per_patient) ## fixed number of mutations
y = matrix(NA, nrow = n*num_samples_per_patient, ncol=d)
for(l in 1:(n*num_samples_per_patient)){
  ## for each patient-specific subsample
  alpha_l = alpha_mat_softmax[l,]*exp(log_lambda_vec_per_obs[l])
  y[l,] = HMP::Dirichlet.multinomial(Nrs = Nm[l], shape = alpha_l)
  
}
y
##-----------------------------------------------------------------------------------------------------##

##-----------------------------------------------------------------------------------------------------##
##' create object with information about covariates, which observations correspond
##' to which patients, and the input (counts)
obj_multilambda <- list(Y = y, x=x, z = z, num_individuals = n)
## obj_multilambda <- readRDS("../data/example_diagDMpatientlambda_dataset.RDS")
# data(obj_multilambda)

##-----------------------------------------------------------------------------------------------------##

##-----------------------------------------------------------------------------------------------------##
## Estimate parameters

## res_patient_lambda <- wrapper_run_TMB(obj_multilambda, model = 'diagDMpatientlambda')
# res_patient_lambda <- readRDS("../data/res_patient_lambda.RDS")
data(res_patient_lambda)
##-----------------------------------------------------------------------------------------------------##


##-----------------------------------------------------------------------------------------------------##
## compare the parameters to the parameters used for the simulation 
fixed_effects_estimate <- matrix(python_like_select_name(res_patient_lambda$par.fixed, 'beta'), nrow=2)
fixed_effects

plot(as.vector(fixed_effects), as.vector(fixed_effects_estimate))
plot((fixed_effects[2,]), (fixed_effects_estimate[2,]))

random_effects_estimate <- matrix(res_patient_lambda$par.random, ncol=(d-1))
plot(as.vector(random_effects), as.vector(random_effects_estimate))

log_lambda_vec_estimate <- python_like_select_name(res_patient_lambda$par.fixed, 'log_lambda')
plot(as.vector(log_lambda_vec), as.vector(log_lambda_vec_estimate))

list_comparisons_true_and_estimate = list(
  all_betas = data.frame(x = as.vector(fixed_effects),
                         y  = as.vector(fixed_effects_estimate)),
  beta_1 =  data.frame(x = fixed_effects[2,],
                       y  = fixed_effects_estimate[2,]),
  random_intercepts = data.frame(x = as.vector(random_effects), 
                                 y = as.vector(random_effects_estimate)),
  log_lambda =  data.frame(x = as.vector(log_lambda_vec), 
                           y = as.vector(log_lambda_vec_estimate)))

plots_comparisons_true_and_estimate <- lapply(1:length(list_comparisons_true_and_estimate), function(i){
  ggplot(list_comparisons_true_and_estimate[[i]], aes(x=x, y=y))+
    geom_point()+
    geom_abline(slope = 1, intercept = 0, lty='dashed')+
    ggtitle(gsub('_', ' ', names(list_comparisons_true_and_estimate)[i]))+
    labs(x='True value', y='Estimate')
})
do.call('grid.arrange', plots_comparisons_true_and_estimate)
```