---
title: "Vignette of the package `CompSign`"
author: "Lena Morrill Gavarró, Dominique-Laurent Couturier, Florian Markowetz"
output: rmarkdown::github_document
vignette: >
  %\VignetteIndexEntry{my-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: "2024"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  cache = FALSE,
  comment = "#>"
)
```

`CompSign` is a toolkit for differential abundance analysis of mutational signatures using a mixed effects Dirichlet-multinominal model (or simpler variations). The compositional nature of mutational signature exposures has often been overlooked but has important implications, as the analyses must be done in relative terms.

## Installation 
`CompSign` can be installed as usual from github:

```{r include=TRUE, eval=FALSE}
library(devtools)
devtools::install_github("lm687/CompSign")
```

These are the libraries used in this vignette:
```{r libraries, include=TRUE, eval=TRUE, message=FALSE}
library(CompSign)
library(gridExtra)
library(TMB)
library(HMP)
library(gridExtra)
library(ggplot2)
theme_set(theme_bw())
```

## Datasets
Several datasets and objects can be loaded as follows using the `data()` function:

```{r, load_data}
data(PancEndocrine_signaturesMSE)
data(ProstAdenoCA_chrom)
data(res_patient_lambda)
data(simplified_object)
```

The package contains the following datasets of exposures of mutational signatures and metadata of the corresponding samples. These datasets are:

- `PancEndocrine_signaturesMSE`: Signature exposures for early and late mutations, in the PCAWG Panc-Endocrine cohort
- `ProstAdenoCA_chrom`: Signature exposures for each chromosome, in the PCAWG Prost-AdenoCA cohort
- `simplified_object`: Nucleotide changes in clonal and subclonal mutations in the Lung-AdenoCA cohort

## Minimal example workflow

A minimal example of the main function `wrapper_run_TMB` is:

```
diagDM_no_small_sigs <- wrapper_run_TMB(object = simplified_object,
                                        model = "diagRE_DM", use_nlminb=T, smart_init_vals=F)
```

in which `simplified_object` is a list containing `simplified_object$x`, `simplified_object$z`, `simplified_object$Y`.

To test for differential abundance, a generalised Wald test can be used with the function `wald_TMB_wrapper`, which gives a p-value as output:

```
wald_TMB_wrapper(diagDM_no_small_sigs)
```

The estimated coefficients can be visualised as follows:
```
library(ggplot2)
plot_betas(diagDM_no_small_sigs, return_ggplot = T)
```


## Example workflow
In this example workflow we determine differential abundance between clonal and subclonal mutations in Pancreatic neuroendocrine tumors.

### `sign` objects: how to load them
`PancEndocrine_signaturesMSE` is an S4 object of class `sign`. `sign` objects include all information about the name of the samples, the mutation signature exposures (of all signatures, or only of the subset of active signatures as determined by Alexandrov et al. 2020). However, the use of these S4 objects is not necessary for running the models of differential abundance, which instead take a list as argument for the input data.

```{r loading_panc-endocrine, message=FALSE, results='hide'}
PancEndocrine_signaturesMSE = load_PCAWG("../inst/extdata/roo/Panc-Endocrine_signaturesMSE_ROO.RDS",
                                         read_directly = T,
                                         typedata = "signaturesMSE", override_warning_X_Z = T)
PancEndocrine_signaturesMSE_v2 = load_PCAWG(ct = "Panc-Endocrine", typedata = "signaturesMSE", path_to_data = "../inst/extdata/", load_all_sigs = F, override_warning_X_Z = T)
```

### Visualisation of mutational signature exposures
All samples - clonal and subclonal - sorted by increasing SBS3 exposure:
```{r panc-endocrine2, fig.height=3.5, fig.width=6, message=FALSE, results='hide'}

createBarplot(normalise_rw(non_duplicated_rows(PancEndocrine_signaturesMSE$Y)),
              order_labels = names(sort(non_duplicated_rows(PancEndocrine_signaturesMSE$Y)[,'SBS3'],
                                        decreasing = F)), remove_labels=T,
              custom_color_palette=color_list)+ggtitle('Sorted by SBS3')

```

### Subsetting the exposures

We create a simplified object containing exposures of fewer signatures (i.e. a subcomposition of the original signature vectors):
```{r simplify_object}
simplified_object <- give_subset_sigs_TMBobj(PancEndocrine_signaturesMSE,
                                             sigs_to_remove = c('SBS13', 'SBS17a', 'SBS17b', 'SBS30'))
```

In which `simplified_object` is a list containing `simplified_object$x`, `simplified_object$z`, `simplified_object$Y`, `simplified_object$d` (see *Input dataset* section below)

The clonal and subclonal exposures are, respectively, the two barplots below:
```{r simplify_object2, fig.height=3, fig.width=7, message=FALSE}
do.call('grid.arrange', list(grobs=lapply(split_matrix_in_half(simplified_object$Y), function(i) createBarplot(normalise_rw(i), remove_labels = T, custom_color_palette=color_list)), nrow=1))
```

## Running the model for differential abundance

### Variations of the model

There are several variations on the model, each with a particular combination of the following:
- Base model: Dirichlet-multinomial in most cases, but two simpler models -- multinomial models `diagRE_M` and `fullRE_M` are implemented for comparison.
- Random effects: correlated or not (e.g. uncorrelated in `diagRE_DM`, correlated in `fullRE_DM`).
- Number of dispersion parameters: generally one per group (e.g. in `diagRE_DM`), but possibly fewer or more if specified in the name (e.g. one single dipersion parameter in `diagREDMsinglelambda`, one dispersion parameter per patient in `diagDMpatientlambda`).

| name of model (for user) | description  |cpp file (no need to use)   |
|---|---|---|
| diagREDMsinglelambda  | DM with independent RE and one lambda  | diagRE_dirichletmultinomial_single_lambda  |
| diagRE_DM  | DM with independent RE and two lambdas  | diagRE_ME_dirichletmultinomial  |
| diagRE_M  | M with independent RE  | diagRE_ME_multinomial  |
| FEDMsinglelambda  | DM with no RE and one lambda  | FE_dirichletmultinomial_single_lambda  |
| FE_DM  | DM with no RE and two lambdas  | FE_dirichletmultinomial  |
| fullREDMsinglelambda  | DM with independent RE and two lambdas  | fullRE_dirichletmultinomial_single_lambda  |
| fullRE_DMonefixedlambda  | DM assuming that there is no overdispersion in the first group (fixed lambda=1)  | fullRE_ME_dirichletmultinomial_onefixedlambda  |
| fullRE_DM  | DM with correlated RE and two lambdas  | fullRE_ME_dirichletmultinomial  |
| fullRE_M  | M with correlated RE  | fullRE_ME_multinomial  |
| singleRE_DM  | DM with a single RE intercept and two lambdas  | singleRE_dirichlet_multinomial  |
| diagDMpatientlambda  | DM with independent RE and one lambda for each patient  | diagREpatientlambda_ME_dirichletmultinomial  |
| fullDMpatientlambda  | DM with correlated RE and one lambda for each patient  | fullREpatientlambda_ME_dirichletmultinomial  |

### Input dataset

The input dataset is the argument `object`, which is a list with the following structure:

- `x`: covariate matrix (`p x n`)
- `z`: matrix of random effects indicating which patient-specific subsample belongs to which patient (`n x N`)
- `Y` (`n x d`)
- `d`

### Minimal example

The first row is the `<model>` argument in the function `wrapper_run_TMB()`.

We run the model `diagREDMsinglelambda` with the dataset `simplified_object`:

```{r panc-endocrine3, message=FALSE, echo=TRUE, results='hide'}
diagDM_no_small_sigs <- wrapper_run_TMB(object = simplified_object,
                                        model = "diagREDMsinglelambda", use_nlminb=T, smart_init_vals=F)
```

This is the resulting object with the estimates and their standard deviations:

```{r TMB_output_example}
diagDM_no_small_sigs
```

We can see how in this case we only have a single estimate for the dispersion parameter (`log_lambda`). The number of $\beta$ (20) corresonds to $2*(d-1)$ where $d$ is the number of mutations in the object (11, as seen in the bar plots above). The last signature or category, SBS39, is used as baseline signature in the ALR transformation. All beta parameters are in ALR-transformed space, using SBS39 as baseline. $\beta_0$ and $\beta_1$ results are intercalated, i.e. the $\beta$ in the first row corresponds to the $\beta_0$ of SBS1 wrt (with respect to) SBS39, the $\beta$ in the third row corresponds to the $\beta_0$ of SBS2 wrt (with respect to) SBS39, the $\beta$ in the first row corresponds to the $\beta_0$ of SBS1 wrt (with respect to) SBS39, etc.

These $\beta$ estimates can be plotted as follows:
```{r plot_betas_example, results='hide', fig.height=2.5, fig.width=5}
plot_betas(diagDM_no_small_sigs)
```

and the p-value indicating differential abundance can be extracted:

```{r, wald_test}
wald_TMB_wrapper(diagDM_no_small_sigs)
```

## Other models
Other models can be run similarly:

```{r, eval=FALSE, echo=TRUE}
res <- wrapper_run_TMB(object = simplified_object,
                       model = "diagREDMsinglelambda", use_nlminb=T, smart_init_vals=F)
res <- wrapper_run_TMB(object = simplified_object,
                       model = "diagRE_DM", use_nlminb=T, smart_init_vals=F)
res <- wrapper_run_TMB(object = simplified_object,
                       model = "diagRE_M", use_nlminb=T, smart_init_vals=F)
res <- wrapper_run_TMB(object = simplified_object,
                       model = "FEDMsinglelambda", use_nlminb=T, smart_init_vals=F)
res <- wrapper_run_TMB(object = simplified_object,
                       model = "FE_DM", use_nlminb=T, smart_init_vals=F)
res <- wrapper_run_TMB(object = simplified_object,
                       model = "fullREDMsinglelambda", use_nlminb=T, smart_init_vals=F)
res <- wrapper_run_TMB(object = simplified_object,
                       model = "fullRE_DMonefixedlambda", use_nlminb=T, smart_init_vals=F)
res <- wrapper_run_TMB(object = simplified_object,
                       model = "fullRE_DM", use_nlminb=T, smart_init_vals=F)
res <- wrapper_run_TMB(object = simplified_object,
                       model = "fullRE_M", use_nlminb=T, smart_init_vals=F)
res <- wrapper_run_TMB(object = simplified_object,
                       model = "singleRE_DM", use_nlminb=T, smart_init_vals=F)
```

## Patient-specific dispersion parameters
In a scenario where we have several samples for each patient, which can still be grouped in two or more groups (fixed effects), we might be interested in patient-specific dispersion parameters, besides the patient-specific intercepts that we had. This has been implemented:

### Simulating data

```{r, patient_dispersion, eval=FALSE}

n <- 40 ## patients
num_samples_per_patient = 30
stopifnot((num_samples_per_patient %% 2)==0)
z <- do.call('rbind', replicate(expr = diag(n), n= num_samples_per_patient, simplify = F))
x <- cbind(rep(1, n*num_samples_per_patient),
           c(rep(0, (n*num_samples_per_patient)/2),
             rep(1, (n*num_samples_per_patient)/2)))
d = 5 ## number of signatures

random_effects <- matrix(runif(n*(d-1)), nrow=n, ncol=d-1)
fixed_effects <- matrix(runif(2*(d-1)), nrow=2, ncol=d-1)

alpha_mat <- x %*% fixed_effects + z %*% random_effects
alpha_mat_softmax = softmax( cbind(alpha_mat, 0) )

## lambda per patient
log_lambda_vec = runif(n, 3, 4)

## get patient-specific overdispersions
log_lambda_vec_per_obs <- z %*% log_lambda_vec

## simulate counts
Nm = rep(1000, n*num_samples_per_patient) ## fixed number of mutations
y = matrix(NA, nrow = n*num_samples_per_patient, ncol=d)
for(l in 1:(n*num_samples_per_patient)){
  ## for each patient-specific subsample
  alpha_l = alpha_mat_softmax[l,]*exp(log_lambda_vec_per_obs[l])
  y[l,] = HMP::Dirichlet.multinomial(Nrs = Nm[l], shape = alpha_l)
  
}

```

```{r, patient_dispersion_inference_list, eval=FALSE}
##' create object with information about covariates, which observations correspond
##' to which patients, and the input (counts)
obj_multilambda <- list(Y = y, x=x, z = z, num_individuals = n)
```

## Estimate parameters
```{r, patient_dispersion_inference, eval=FALSE}
res_patient_lambda <- wrapper_run_TMB(obj_multilambda, model = 'diagDMpatientlambda')
```

```{r, patient_dispersion_inference_res}
data(res_patient_lambda)
data(obj_multilambda)
data(obj_multilambda_parameters)
```


## Compare the estimates to the ground truth

The true parameters used for the simulation are found in the object `obj_multilambda_parameters`.

We can see how the parameters of interest are well recovered:

  - $\beta_1$
  - $\log(\lambda)$

and two additional sets of parameters suffer from a bias due to having used the uncorrelated version of the model (i.e. `diagDMpatientlambda` instead of `fullDMpatientlambda`):

  - $\beta_0$
  - Random intercepts for each patient
Note that in the current setting the number of parameters to estimate is already large, and this is why `diagDMpatientlambda` is used.

```{r, include=F, eval=TRUE}
appender <- function(string){
    sapply(string, function(stringb){
      if(stringb == 'Intercept'){
        TeX(paste("$\\beta_0$")) 
      }else if (stringb == 'Slope'){
        TeX(paste("$\\beta_1$")) 
      }
    })
}
```

```{r, patient_dispersion_inference_analysis, fig.width=10}
fixed_effects_estimate <- matrix(python_like_select_name(res_patient_lambda$par.fixed, 'beta'), nrow=2)
random_effects_estimate <- matrix(res_patient_lambda$par.random, ncol=ncol(obj_multilambda_parameters$fixed_effects))
log_lambda_vec_estimate <- python_like_select_name(res_patient_lambda$par.fixed, 'log_lambda')

list_comparisons_true_and_estimate = list(
  all_betas = data.frame(x = as.vector(obj_multilambda_parameters$fixed_effects),
                         y  = as.vector(fixed_effects_estimate)),
  beta_1 =  data.frame(x = obj_multilambda_parameters$fixed_effects[2,],
                       y  = fixed_effects_estimate[2,]),
  random_intercepts = data.frame(x = as.vector(obj_multilambda_parameters$random_effects), 
                                 y = as.vector(random_effects_estimate)),
  log_lambda =  data.frame(x = as.vector(obj_multilambda_parameters$log_lambda_vec), 
                           y = as.vector(log_lambda_vec_estimate)))

plots_comparisons_true_and_estimate <- lapply(1:length(list_comparisons_true_and_estimate), function(i){
  ggplot(list_comparisons_true_and_estimate[[i]], aes(x=x, y=y))+
    geom_point()+
    geom_abline(slope = 1, intercept = 0, lty='dashed')+
    ggtitle(gsub('_', ' ', names(list_comparisons_true_and_estimate)[i]))+
    labs(x='True value', y='Estimate')
})
do.call('grid.arrange', c(plots_comparisons_true_and_estimate, nrow=1))
```

