---
title: "Vignette of the package `CompSign`"
author: "Lena Morrill Gavarró, Dominique-Laurent Couturier, Florian Markowetz"
output: rmarkdown::github_document
vignette: >
  %\VignetteIndexEntry{my-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: "2024"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  cache = FALSE,
  comment = "#>"
)
```

`CompSign` is a toolkit for differential abundance analysis of mutational signatures using a mixed effects Dirichlet-multinominal model (or simpler variations). The compositional nature of mutational signature exposures has often been overlooked but has important implications, as the analyses must be done in relative terms.

## Installation 
`CompSign` can be installed as usual from github:

```{r include=TRUE, eval=FALSE}
library(devtools)
devtools::install_github("lm687/CompSign")
```

These are the libraries used in this vignette:
```{r libraries, include=TRUE, eval=TRUE, message=FALSE}
library(CompSign)
library(gridExtra)
library(TMB)
library(HMP)
library(gridExtra)
library(ggplot2)
theme_set(theme_bw())
```

## Datasets
Several datasets and objects can be loaded as follows using the `data()` function:

```{r, load_data}
data(PancEndocrine_signaturesMSE)
data(ProstAdenoCA_chrom)
data(res_patient_lambda)
data(simplified_object)
data(ProstAdenoCA_trinucleotide)
data(obj_multilambda)
data(obj_multilambda_parameters)
```

The package contains the following datasets of exposures of mutational signatures and metadata of the corresponding samples. These datasets are:

- `PancEndocrine_signaturesMSE`: Signature exposures for early and late mutations, in the PCAWG Panc-Endocrine cohort
- `ProstAdenoCA_chrom`: Signature exposures for each chromosome, in the PCAWG Prost-AdenoCA cohort
- `simplified_object`: Nucleotide changes in clonal and subclonal mutations in the Lung-AdenoCA cohort
- `ProstAdenoCA_trinucleotide`: Trinucleotide abundandances for clonal and subclonal mutations in the PCAWG Prost-AdenoCA cohort

## Minimal example workflow (Example 1)

This is a minimal example for running the model and extracting the estimated coefficients.

### Input dataset

The input dataset is the argument `object`, which is a list with the following structure:
- `x`: covariate matrix (`p x n`)
- `z`: matrix of random effects indicating which patient-specific subsample belongs to which patient (`n x N`)
- `Y` (`n x d`)
- `d`

### Fitting the model

The function `wrapper_run_TMB()` is used to run all variations of the model.

A minimal example is:

```
diagDM_no_small_sigs <- wrapper_run_TMB(object = simplified_object,
                                        model = "diagRE_DM", use_nlminb=T, smart_init_vals=F)
```

in which `simplified_object` is a list containing `simplified_object$x`, `simplified_object$z`, `simplified_object$Y`. The `object` input for `wrapper_run_TMB` is always a list containing a matrix of covariates `x`, a matrix of observations `Y` and a matrix matching observations patients and the labels corresponding to the random effects. All three matrices should have the same number of rows - the number of observations. In this case, `simplified_object` corresponds to the nucleotides abundances in clonal and subclonal mutations in the Lung-AdenoCA cohort. `simplified_object$x`, contains two columns (a column of 1 for the intercept and a column of 0 and 1 indicating whether the observation corresponds to clonal (0) or subclonal (1) mutations. `simplified_object$z` contains a matrix where each observation is paired with its corresponding patient, `simplified_object$Y` is a matrix of counts for the nucleotide abundances for each patient and (clonal/subclonal) group.

Here we are using the model `diagRE_DM`, which is a mixed-effect Dirichlet-multinomial model with non-correlated random effects and two dispersion parameters (one per group). See below for the list of alternative models available.

### Data included in CompSign
You can load `simplified_object`, as well as several other example datasets, using the `data` funtion:

```
data(package='CompSign')
data(simplified_object)
```

### Testing for differential abundance

Depending on the model, the output of `wrapper_run_TMB` contains the estimated values of different coefficients. The parameters of interest are `beta_0`, `beta_1` and `lambda`. Both `beta` are values in log-ratio space in with respect to the last category included in the input `object` for `wrapper_run_TMB` (i.e. the last colun of `object$Y`). `beta_0` indicates the coefficient for the baseline - in the two-group example of `simplified_object`, this baseline reflects the abundance in the first (clonal) group. `beta_1` indicates the cofficient for the second column - here representing the difference between clonal and subclonal mutations. If `beta_1` is a vector of 0, this indicates that none of the log-ratios of abundances have changed between the two groups, and hence that there is no differential abundance.

To test for differential abundance, a generalised Wald test can be used with the function `wald_TMB_wrapper`, which gives a p-value as output:

```
wald_TMB_wrapper(diagDM_no_small_sigs)
```

### Visualising the output

The estimated `beta` coefficients can be visualised as follows:

```
library(ggplot2)
plot_betas(diagDM_no_small_sigs, return_ggplot = T)
```


## Workflow with mutational signatures (example 2)
In this example workflow we determine differential abundance between clonal and subclonal mutations in Pancreatic neuroendocrine tumors.

### `sign` objects: how to load them
`PancEndocrine_signaturesMSE` is an S4 object of class `sign`. `sign` objects include all information about the name of the samples, the mutation signature exposures (of all signatures, or only of the subset of active signatures as determined by Alexandrov et al. 2020). However, the use of these S4 objects is not necessary for running the models of differential abundance, which instead take a list as argument for the input data.

```{r loading_panc-endocrine, message=FALSE, results='hide'}
PancEndocrine_signaturesMSE = load_PCAWG("../inst/extdata/roo/Panc-Endocrine_signaturesMSE_ROO.RDS",
                                         read_directly = T,
                                         typedata = "signaturesMSE", override_warning_X_Z = T)
PancEndocrine_signaturesMSE_v2 = load_PCAWG(ct = "Panc-Endocrine", typedata = "signaturesMSE", path_to_data = "../inst/extdata/", load_all_sigs = F, override_warning_X_Z = T)
```

### Signature extraction
If signature extraction has not been performed, the wrapper function `extract_sigs_TMB_obj` outputs signature exposures from a matrix of trinucleotide abundances. We use the trinucleotide abundances from the object `ProstAdenoCA_trinucleotide`. In this wrapper function two methods for signature extraction are implemented: mutSigExtractor and quadratic programming (QP); this can be specified in the `signature_fitting_method` argument. The version of signature definitions matrix can be specified in the argument `signature_version` (which can be `v2` or `v3` of the COSMIC signatures. Alternatively, any signature definition matrix can be specified in the `signature_definition`, where the signature definition should be a `F x K` signature definition matrix, where `K` is the number of signatures and `F` the number of features (for COSMIC signatures, `F=96`). The active signature list can be specified in `subset_signatures`.

```{r signature_extraction}
active_sigs_in_ct <- c("SBS1", "SBS2", "SBS3", "SBS5", "SBS8", "SBS13", "SBS18", "SBS37", "SBS40", "SBS41")
exposures_from_mutSigExtractor <- extract_sigs_TMB_obj(dataset_obj_trinucleotide=ProstAdenoCA_trinucleotide,
                             subset_signatures = active_sigs_in_ct,
                             signature_version=NULL, signature_fitting_method = 'mutSigExtractor')

exposures_from_QP <- extract_sigs_TMB_obj(dataset_obj_trinucleotide=ProstAdenoCA_trinucleotide,
                              subset_signatures =active_sigs_in_ct,
                              signature_version=NULL, signature_fitting_method = 'QP')

plot(log(as.vector(exposures_from_mutSigExtractor$Y)), log(as.vector(exposures_from_QP$Y)))
```

### Visualisation of mutational signature exposures
All samples - clonal and subclonal - sorted by increasing SBS3 exposure:
```{r panc-endocrine2, fig.height=3.5, fig.width=6, message=FALSE, results='hide'}

createBarplot(normalise_rw(non_duplicated_rows(PancEndocrine_signaturesMSE$Y)),
              order_labels = names(sort(non_duplicated_rows(PancEndocrine_signaturesMSE$Y)[,'SBS3'],
                                        decreasing = F)), remove_labels=T,
              custom_color_palette=color_list)+ggtitle('Sorted by SBS3')

```

### Subsetting the exposures

We create a simplified object containing exposures of fewer signatures (i.e. a subcomposition of the original signature vectors):
```{r simplify_object}
simplified_object <- give_subset_sigs_TMBobj(PancEndocrine_signaturesMSE,
                                             sigs_to_remove = c('SBS13', 'SBS17a', 'SBS17b', 'SBS30'))
```

In which `simplified_object` is a list containing `simplified_object$x`, `simplified_object$z`, `simplified_object$Y`, `simplified_object$d` (see *Input dataset* section below)

The clonal and subclonal exposures are, respectively, the two barplots below:
```{r simplify_object2, fig.height=3, fig.width=7, message=FALSE}
do.call('grid.arrange', list(grobs=lapply(split_matrix_in_half(simplified_object$Y), function(i) createBarplot(normalise_rw(i), remove_labels = T, custom_color_palette=color_list)), nrow=1))
```

### Variations of the model

There are several variations on the model, each with a particular combination of the following:
- Base model: Dirichlet-multinomial in most cases, but two simpler models -- multinomial models `diagRE_M` and `fullRE_M` are implemented for comparison.
- Random effects: correlated or not (e.g. uncorrelated in `diagRE_DM`, correlated in `fullRE_DM`).
- Number of dispersion parameters: generally one per group (e.g. in `diagRE_DM`), but possibly fewer or more if specified in the name (e.g. one single dipersion parameter in `diagRE_DM_singlelambda`, one dispersion parameter per patient in `diagRE_DM_patientlambda`).

The names of the model correspond to `{random effects}_{base model}_{particularities of the dispersion parameter, if any}`.

The first column is the `<model>` argument in the function `wrapper_run_TMB()`.
| Name of model (for user) | Description and usage  |
|---|---|
| `FE_DM_singlelambda`  | Dirichlet-multinomial  with no RE and a single `lambda` which can be used in a two-group comparison (if we consider the dipersion to be the same in both groups), a multi-group comparison, or any regression setting.  | 
| `FE_DM`  | Dirichlet-multinomial  with no random effects and two `lambda`, used for the comparison between two groups in which we want to account for group-specific dispersion. Model slightly more complex than `FE_DM_singlelambda` | 
| `diagRE_M`  |  Multinomial with non-correlated multivariate effects. Simpler model than `diagRE_DM`, as no dispersion is included | 
| `diagRE_DM_singlelambda`  |  Dirichlet-multinomial with non-correlated multivariate RE and one: equivalent of diagRE_DM but with a shared . It can be used in a two-group comparison (if we consider the dipersion to be the same in both groups), a multi-group comparison, or any regression setting.  | 
| `singleRE_DM`  | Dirichlet-multinomial with a single RE intercept and two `lambda`: simple model in which random effects are not multivariate, but where group-specific dispersion is needed, in a two-group comparison. | 
| `diagRE_DM`  |  Dirichlet-multinomial with independent RE and two `lambda`: used most commonly throughout the paper. The data are matched to warrant multivariate RE, but correlations between categories are not explicitly modelled. Faster than `fullRE_DM` with often comparable results for `beta_1`.   | 
| `fullRE_DM_singlelambda`  | Dirichlet-multinomial with correlated RE and two lambdab: equivalent of `diagR_DM_singlelambda` that can be used if categories have strong correlations. | 
| `fullRE_M`  | Multinomial with correlated RE: assuming no overdispersion, it can be used in any regression setting  | 
| `fullRE_DM`  | Dirichlet-multinomial with correlated RE and two `lambda`: model with multivariate RE with correlations modelled explicitly. As a `(K-1)*(K-1)` covariance matrix is estimated, this model is not recommended where the ratio of signatures to samples is high.  |
| `diagRE_DM_patientlambda`  | Dirichlet-multinomial with non-correlated RE and one lambda$ per patient: model more complex than `diag_RE_DM` that can be used when there are several observations per patient and we wish to include a patient-specific dispersion parameter  | 
| `fullRE_DM_patientlambda`  |  Dirichlet-multinomial with correlated RE and one  per patient: equivalent of `diagRE_DM_patientlambda`, but modelling correlations between categories | 


### Using an alternative model, and parameter interpretation (example 3)

We run the model `diagRE_DM_singlelambda` with the dataset `simplified_object`:

```{r panc-endocrine3, message=FALSE, echo=TRUE, results='hide'}
diagDM_no_small_sigs <- wrapper_run_TMB(object = simplified_object,
                                        model = "diagRE_DM_singlelambda", use_nlminb=T, smart_init_vals=F)
```

This is the resulting object with the estimates and their standard deviations:

```{r TMB_output_example}
diagDM_no_small_sigs
```

We can see how in this case we only have a single estimate for the dispersion parameter (`log_lambda`). The number of $\beta$ (20) corresonds to $2*(d-1)$ where $d$ is the number of mutations in the object (11, as seen in the bar plots above). The last signature or category, SBS39, is used as baseline signature in the ALR transformation. All beta parameters are in ALR-transformed space, using SBS39 as baseline. $\beta_0$ and $\beta_1$ results are intercalated, i.e. the $\beta$ in the first row corresponds to the $\beta_0$ of SBS1 wrt (with respect to) SBS39, the $\beta$ in the third row corresponds to the $\beta_0$ of SBS2 wrt (with respect to) SBS39, the $\beta$ in the first row corresponds to the $\beta_0$ of SBS1 wrt (with respect to) SBS39, etc.

These $\beta$ estimates can be plotted as follows:
```{r plot_betas_example, results='hide', fig.height=2.5, fig.width=5}
plot_betas(diagDM_no_small_sigs)
```

## Using an alternative model with patient-specific dispersion (example 4)
In a scenario where we have several samples for each patient, which can still be grouped in two or more groups (fixed effects), we might be interested in patient-specific dispersion parameters, besides the patient-specific intercepts that we had. This has been implemented:

### Simulating data

```{r, patient_dispersion, eval=FALSE}

n <- 40 ## patients
num_samples_per_patient = 30
stopifnot((num_samples_per_patient %% 2)==0)
z <- do.call('rbind', replicate(expr = diag(n), n= num_samples_per_patient, simplify = F))
x <- cbind(rep(1, n*num_samples_per_patient),
           c(rep(0, (n*num_samples_per_patient)/2),
             rep(1, (n*num_samples_per_patient)/2)))
d = 5 ## number of signatures

random_effects <- matrix(runif(n*(d-1)), nrow=n, ncol=d-1)
fixed_effects <- matrix(runif(2*(d-1)), nrow=2, ncol=d-1)

alpha_mat <- x %*% fixed_effects + z %*% random_effects
alpha_mat_softmax = softmax( cbind(alpha_mat, 0) )

## lambda per patient
log_lambda_vec = runif(n, 3, 4)

## get patient-specific overdispersions
log_lambda_vec_per_obs <- z %*% log_lambda_vec

## simulate counts
Nm = rep(1000, n*num_samples_per_patient) ## fixed number of mutations
y = matrix(NA, nrow = n*num_samples_per_patient, ncol=d)
for(l in 1:(n*num_samples_per_patient)){
  ## for each patient-specific subsample
  alpha_l = alpha_mat_softmax[l,]*exp(log_lambda_vec_per_obs[l])
  y[l,] = HMP::Dirichlet.multinomial(Nrs = Nm[l], shape = alpha_l)
  
}

```

```{r, patient_dispersion_inference_list, eval=FALSE}
##' create object with information about covariates, which observations correspond
##' to which patients, and the input (counts)
obj_multilambda <- list(Y = y, x=x, z = z, num_individuals = n)
```

## Estimate parameters
```{r, patient_dispersion_inference, eval=FALSE}
res_patient_lambda <- wrapper_run_TMB(obj_multilambda, model = 'diagRE_DM_patientlambda')
```

```{r, patient_dispersion_inference_res}
data(res_patient_lambda)
data(obj_multilambda)
data(obj_multilambda_parameters)
```


### Comparison of estimates with ground truth

The true parameters used for the simulation are found in the object `obj_multilambda_parameters`.

We can see how the parameters of interest are well recovered:

  - $\beta_1$
  - $\log(\lambda)$

and two additional sets of parameters suffer from a bias due to having used the uncorrelated version of the model (i.e. `diagRE_DM_patientlambda` instead of `fullDMpatientlambda`):

  - $\beta_0$
  - Random intercepts for each patient
Note that in the current setting the number of parameters to estimate is already large, and this is why `diagRE_DM_patientlambda` is used.

```{r, include=F, eval=TRUE}
appender <- function(string){
    sapply(string, function(stringb){
      if(stringb == 'Intercept'){
        TeX(paste("$\\beta_0$")) 
      }else if (stringb == 'Slope'){
        TeX(paste("$\\beta_1$")) 
      }
    })
}
```

```{r, patient_dispersion_inference_analysis, fig.width=10}
fixed_effects_estimate <- matrix(python_like_select_name(res_patient_lambda$par.fixed, 'beta'), nrow=2)
random_effects_estimate <- matrix(res_patient_lambda$par.random, ncol=ncol(obj_multilambda_parameters$fixed_effects))
log_lambda_vec_estimate <- python_like_select_name(res_patient_lambda$par.fixed, 'log_lambda')

list_comparisons_true_and_estimate = list(
  all_betas = data.frame(x = as.vector(obj_multilambda_parameters$fixed_effects),
                         y  = as.vector(fixed_effects_estimate)),
  beta_1 =  data.frame(x = obj_multilambda_parameters$fixed_effects[2,],
                       y  = fixed_effects_estimate[2,]),
  random_intercepts = data.frame(x = as.vector(obj_multilambda_parameters$random_effects), 
                                 y = as.vector(random_effects_estimate)),
  log_lambda =  data.frame(x = as.vector(obj_multilambda_parameters$log_lambda_vec), 
                           y = as.vector(log_lambda_vec_estimate)))

plots_comparisons_true_and_estimate <- lapply(1:length(list_comparisons_true_and_estimate), function(i){
  ggplot(list_comparisons_true_and_estimate[[i]], aes(x=x, y=y))+
    geom_point()+
    geom_abline(slope = 1, intercept = 0, lty='dashed')+
    ggtitle(gsub('_', ' ', names(list_comparisons_true_and_estimate)[i]))+
    labs(x='True value', y='Estimate')
})
do.call('grid.arrange', c(plots_comparisons_true_and_estimate, nrow=1))
```

